---
title: "Employee Attrition and Annual Leave Predictor"
author: "Venura Gunaratne"
editor: visual
date: today
format: 
  typst: 
    toc: true
    number-sections: true
    columns: 1
    margin: 
      top: 2.5cm
      bottom: 2.5cm
      left: 2cm
      right: 2cm
    header: Attrition and Annual Leave Predictor Engine
    footer: |
      Venura Gunaratne
      hfil
      Page 并 
---

```{r setup, include=FALSE}
library(tidyverse)
library(knitr)
```

## The logistic regression model highlights key drivers of employee attrition expressed as Odds Ratios (OR):

```{r odds-ratio-table, echo=FALSE}

read_csv("outputs/attrition_odds_ratios.csv", show_col_types = FALSE) |>
  filter(term != "(Intercept)") |>
  select(Term = term, 'Odds Ratio' = estimate, '2.5% CI' = conf.low, '97.5% CI' = conf.high, 'pvalue' = p.value) |>
  mutate(across(where(is.numeric), ~round(.x, 3))) |>
  kable(caption = "Table 1: Attrtion Odds Ratios with 95% Confidence Intervals")
```

## Predictors for annual leave frequency modeled via Poisson regression (Incident Rate Ratios):

```{r irr-table, echo=FALSE}

read_csv("outputs/leave_irr.csv", show_col_types = FALSE) |>
  filter(term != "(Intercept)") |>
  select(Term = term, 'IRR' = estimate, '2.5% CI' = conf.low, '97.5% CI' = conf.high, 'pvalue' = p.value) |>
  mutate(across(where(is.numeric), ~round(.x, 3))) |>
  kable(caption = "Table 2: Leave Frequency Incident Rate Ratios with 95% Confidence Intervals")
```

## RMSE and MAE

```{r poisson-metrics, echo=FALSE, fig.cap="Out-of-Sample Performance Metrics for Poisson Leave Prediction Engine"}

library(tidyverse)

# Read the Poisson metrics CSV
poisson_metrics_df <- read_csv("outputs/poisson_metrics.csv", show_col_types = FALSE) |> 
  mutate(
    metric = toupper(metric),
    metric = factor(metric, levels = c("RMSE", "MAE"))
  )

# Plot RMSE and MAE
ggplot(poisson_metrics_df, aes(x = metric, y = value, fill = metric)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = round(value, 3)), vjust = -0.5, fontface = "bold", size = 4) +
  scale_fill_manual(values = c("RMSE" = "#2c3e50", "MAE" = "#e74c3c")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Poisson Engine Out-of-Sample Error Metrics",
    subtitle = "Lower values indicate higher predictive accuracy on test data",
    x = "Evaluation Metric",
    y = "Error (In Days/Leaves)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray30")
  )
```

## Model Performance & Discrimination

```{r roc-curve-plot, echo=FALSE, fig.cap="Model Performance and ROC-AUC"}
library(tidyverse)

# 1. Read exported metric and ROC curve data
metrics <- read_csv("outputs/model_metrics.csv", show_col_types = FALSE)
roc_coords <- read_csv("outputs/attrition_roc_curve.csv", show_col_types = FALSE)

# 2. Extract the ROC-AUC value dynamically
auc_value <- metrics %>%
  filter(model == "Logistic Regression", metric == "ROC-AUC") %>%
  pull(value) %>%
  round(3)

# 3. Create interactive or static plot with the dynamic AUC value
ggplot(roc_coords, aes(x = 1 - specificity, y = sensitivity)) +
  geom_path(color = "#2b5c8f", linewidth = 1.2) +
  geom_abline(lty = 2, color = "gray50") +
  annotate(
    "text", x = 0.65, y = 0.25, 
    label = paste("Model ROC-AUC =", auc_value), 
    size = 5, fontface = "bold", color = "#2b5c8f"
  ) +
  labs(
    title = "Attrition Model ROC Curve",
    subtitle = paste("Evaluated on unseen test set (AUC:", auc_value, ")"),
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)"
  ) +
  theme_minimal()
```

The model achieved an out-of-sample **ROC-AUC of 0.837**, demonstrating strong predictive capability in distinguishing high-risk attrition candidates from stable employees on unseen evaluation data.

## Top 10 Predictors of Attrition

```{r vip-plot, echo=FALSE, fig.cap="Top 10 Feature Importance Scores for Attrition Prediction"}

library(tidyverse)

# Read the exported VIP data
vip_data <- read_csv("outputs/attrition_vip.csv", show_col_types = FALSE)

# Plot using standard ggplot2
ggplot(vip_data, aes(x = reorder(term, importance), y = importance)) +
  geom_col(fill = "#2b5c8f") +
  coord_flip() +
  labs(
    title = "Top 10 Predictors Driving Attrition",
    x = "Predictor Term",
    y = "Importance (|z-score|)"
  ) +
  theme_minimal()
```

## Comparison Between Log Regression and XGBoost Models

```{r model-comparison-table, echo=FALSE}

library(tidyverse)
library(knitr)

# Read model metrics CSV
metrics_table <- read_csv("outputs/model_metrics.csv", show_col_types = FALSE)

# Format table using knitr
kable(
  metrics_table, 
  digits = 3, 
  col.names = c("Model Architecture", "Performance Metric", "Test Score"),
  caption = "Out-of-Sample Performance Comparison"
)
```

Both baseline Logistic Regression and hyperparameter-tuned XGBoost were evaluated on an untouched 20% test data set (stratified by attrition status).

## Comparative ROC Curve Between the Two Models

```{r combined-roc-plot, echo=FALSE, fig.cap="Combined ROC Curves for Attrition Prediction Models"}

library(tidyverse)

# Read combined ROC curve coordinates
roc_data <- read_csv("outputs/combined_roc_curves.csv", show_col_types = FALSE)

# Fetch latest AUC scores dynamically for the plot subtitle
auc_scores <- read_csv("outputs/model_metrics.csv", show_col_types = FALSE)
log_auc <- round(auc_scores$value[auc_scores$model == "Logistic Regression"], 3)
xgb_auc <- round(auc_scores$value[auc_scores$model == "XGBoost"], 3)

# Render Comparative ROC Plot
ggplot(roc_data, aes(x = 1 - specificity, y = sensitivity, color = model)) +
  geom_path(linewidth = 1.2) +
  geom_abline(lty = 2, color = "gray50", intercept = 0, slope = 1) +
  scale_color_manual(values = c("Logistic Regression" = "#2b5c8f", "XGBoost" = "#d95f02")) +
  labs(
    title = "ROC Curve Comparison (Logistic Regression vs. XGBoost)",
    subtitle = paste0("Logistic Regression AUC: ", log_auc, " | XGBoost AUC: ", xgb_auc),
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)",
    color = "Model Architecture"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Combined Multi-Model Comparison

```{r tbl-all-models-comparison, echo=FALSE, tbl_cap="Comprehensive Predictive Engine Comparison Across Attrition and Leave Outcomes"}

library(tidyverse)
library(knitr)

# Read combined model metrics CSV
all_metrics_df <- read_csv("outputs/all_models_metrics.csv", show_col_types = FALSE)

# Format as a Quarto Markdown Table
all_metrics_df |> 
  rename(
    `Model Engine` = model,
    `Target Outcome` = target,
    `Metric` = metric,
    `Test Score` = value
  ) |> 
  kable(digits = 3, align = "lllc")
```

```{r fig-all-model-chart, echo=FALSE, fig.cap="Comparative Overview of Model Metrics Across Prediction Engines"}

# Visual chart for the combined model metrics
all_metrics_df |> 
  mutate(display_label = paste0(model, "\n(", metric, ")")) |> 
  ggplot(aes(x = display_label, y = value, fill = model)) +
  geom_col(width = 0.55, show.legend = FALSE) +
  geom_text(aes(label = round(value, 3)), vjust = -0.5, fontface = "bold") +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(
    title = "Overall Predictive Engine Metric Comparison",
    subtitle = "ROC-AUC evaluated for Attrition models | RMSE & MAE evaluated for Poisson Leave model",
    x = "Model Engine & Evaluated Metric",
    y = "Metric Score"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray30")
  )
```

## Monte Carlo Model Summary

```{r tbl-monte-carlo-summary, echo=FALSE, tbl_cap="Monte Carlo Workforce Risk Distribution (10,000 Simulated Futures)"}

library(tidyverse)
library(knitr)

#Actual calculated mc_summary.
mc_summary <- read_csv("outputs/monte_carlo_summary.csv", show_col_types = FALSE)

# Construct a 2-row table directly from mc_summary metrics
tibble(
  `Simulated Outcome` = c("Total Resignations (Attrition)", "Total Cumulative Leave Days"),
  `Expected Mean`     = c(round(mc_summary$mean_attrition, 1), round(mc_summary$mean_leaves, 1)),
  `Lower 2.5% Bound`  = c(round(mc_summary$attrition_p2_5, 1), round(mc_summary$leaves_p2_5, 1)),
  `Upper 97.5% Bound` = c(round(mc_summary$attrition_p97_5, 1), round(mc_summary$leaves_p97_5, 1))
) |> 
  kable(align = "lccc")
```

## Monte Carlo Distribution

```{r monte-carlo-distribution, echo=FALSE, fig.cap="Simulated Organizational Attrition Distribution Across 10,000 Monte Carlo Futures"}

simulation_results <- read_csv("outputs/monte_carlo_results.csv", show_col_types = FALSE)

ggplot(simulation_results, aes(x = total_attrition)) +
  geom_histogram(binwidth = 1, fill = "#2c3e50", color = "white", alpha = 0.85) +
  geom_vline(xintercept = mc_summary$mean_attrition, color = "#e74c3c", linewidth = 1.2) +
  geom_vline(xintercept = c(mc_summary$attrition_p2_5, mc_summary$attrition_p97_5), color = "#e67e22", linetype = "dashed", linewidth = 1) +
  annotate("text", x = mc_summary$mean_attrition, y = Inf, label = paste("Mean:", round(mc_summary$mean_attrition, 1)), vjust = 2, color = "#e74c3c", fontface = "bold") +
  labs(
    title = "Monte Carlo Attrition Risk Distribution",
    subtitle = "Dashed orange lines mark the 95% simulated interval bounds",
    x = "Simulated Number of Resignations",
    y = "Frequency across Runs"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```
